---
title: "Project 1 Kiran"
output: html_document
---
```{r}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", 
    warning = F, message = F, tidy = TRUE, tidy.opts = list(width.cutoff = 60), 
    R.options = list(max.print = 100))
```

### 0: Introduction (5pts)
```{r}
library(readr)
hapiscore_whr_2_ <- read_csv("hapiscore_whr.csv")
murder_total_deaths_2_ <- read_csv("murder_total_deaths(2).csv")
income_per_person_gdppercapita_ppp_inflation_adjusted_2_ <-read_csv("income_per_person_gdppercapita_ppp_inflation_adjusted_2.csv")
```
For this project, I chose three different datasets. These datasets were: a country’s happiness score, the total deaths by murder in each country, and the average income per person in each country. I chose these three datasets since I thought it would be interesting to see if there was a correlation between a country’s happiness and their average income or the number of deaths by murder. All three data sets have the two same categorical variables - Country and Year (2005 - 2010). Alongside the categorical variables, the three numerical variables used in this project were each countries happiness rating, the average income per individual, and the total number of deaths by murder in each country. All data used in this project was acquired at www.gapminder.com. I hypothesize that happier countries will have higher average incomes and lower murder rates. 

### 1. Tidying: Rearranging Wide/Long (10 pts)
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
#tidy
tidyhappiness <- hapiscore_whr_2_ %>% gather("2005", 
        "2006", "2007", "2008", "2009", "2010", "2011", "2012", 
        "2013", "2014", "2015", key = "year", 
        value = "happiness scores") %>% arrange(country,year)
tidyincome <- income_per_person_gdppercapita_ppp_inflation_adjusted_2_ %>% gather("2005", 
        "2006", "2007", "2008", "2009", "2010", "2011", "2012", 
        "2013", "2014", "2015", key = "year", 
        value = "income per person") %>% arrange(country,year)
tidymurderdeaths <- murder_total_deaths_2_ %>% gather("2005", 
        "2006", "2007", "2008", "2009", "2010", "2011", "2012", 
        "2013", "2014", "2015", key = "year", 
        value = "deaths by murder") %>% arrange(country,year)
#remove NA values
tidyhappiness<-tidyhappiness %>% na.omit() 
tidyincome <-tidyincome %>% na.omit() 
tidymurderdeaths <-tidymurderdeaths %>% na.omit() 
```
None of my datasets were tidy; therefore, I tidied them using the gather function. By doing so, I made sure that every column was a variable and that every row was an observation. I did so by gathering all years from 2005 - 2015 (since all datasets had these years in common) and created a new column with them titled “Year”. Additionally, I made a new corresponding column for each dataset. For the happiness dataset, I created a new column titled “happiness score,” which ranked each country on a happiness scale from 0 to 1. For the income dataset, I made a new column titled “income per person,” which included all the average incomes per person in each country and year. For my murders dataset, I created a column titled “death by murders,” representing the total number of deaths due to murders in each country every year. Lastly, I removed any NA values from all datasets using the na.omit() function. 

### 2. Joining/Merging (10 pts)
```{r}
library(dplyr)
mergeddata <- tidyhappiness %>% left_join(tidyincome) %>% 
    left_join(tidymurderdeaths)
```
To merge all three of my datasets, I found a common variable between them. In this case, all of my datasets had two common variables, ‘Year’ and ‘Country.’ I was then able to merge them using the left_join function. No cases had to be dropped since each dataset had the same years and countries. After successfully combining my data, the new dataset (called ‘mergeddata’) had 1,251 observations. My three numeric variables used were happiness scores, income, and deaths. 

### 3. Wrangling (40 pts)

DPLYR functions:
```{r}
library(tidyverse)
mergeddata %>% group_by(country) %>% summarize(mean(`deaths by murder`, 
    na.rm = T)) %>% head()
mergeddata %>% select(country,`happiness scores`) %>% head()

mergeddata %>% mutate(`happiness scores (1-10)` = `happiness scores`* 10) %>% head()

mergeddata %>% filter(year == "2015") %>% arrange(desc(`deaths by murder`, 
    na.rm = T))
```

First, I used the ‘group_by’ and the ‘summarize’ function to find the country with the most deaths by murder, which was Brazil which was 61,000 deaths by murder every year. Next, I used the ‘select’ function to select for ‘happiness scores’ and find the country with the highest score: Denmark, with a score of 0.802. To make the data easier to read, I used the ‘mutate’ function to create a new variable called “happiness scores (1-10)”, which converted all the happiness scores into whole numbers from a scale of 1 - 10. Lastly, I used the ‘filter’ function to get all the data from 2015 and used the ‘arrange’ function to organize the number of deaths in descending order. I found that Denmark (who had the highest happiness score) also had one of the lowest death by murder rates. (To save room when knitting the file, I added head() at the end of each line of code.)

Summary Statistics:
```{r}
mergeddata %>% group_by(country) %>% summarize_if(is.numeric,
    mean, na.rm = T) %>% head()
mergeddata %>% group_by(country) %>% summarize_if(is.numeric,
    sd, na.rm = T) %>% head()
mergeddata %>% group_by(country) %>% summarize_if(is.numeric,
    var, na.rm = T) %>% head()
mergeddata %>% group_by(country) %>% summarize_if(is.numeric,
    min, na.rm = T) %>% head()
mergeddata %>% group_by(country) %>% summarize_if(is.numeric,
    max, na.rm = T) %>% head()
```
Using the summary statistics functions, I was able to analyze my numerical variables further. The five summary statistics I used were mean, standards deviation, variation, min, and max. In the first set of codes (up above), I grouped my numerical values by ‘year.’ By doing this, I was able to view the mean, sd, min, and max of happiness scores, incomes, and death rates by year. One interesting result I found was that the countries’ mean happiness scores decreased as the years had progressed, while mean income and deaths by murder increased. In my second set of codes (down below), I grouped my numerical values by ‘country.’ This had allowed me to compare the different numerical data between countries and look for any correlation. 
```{r}
mergeddata %>% group_by(year) %>% summarize_if(is.numeric,
    mean, na.rm = T) %>% head()
mergeddata %>% group_by(year) %>% summarize_if(is.numeric,
    sd, na.rm = T) %>% head()
mergeddata %>% group_by(year) %>% summarize_if(is.numeric,
    var, na.rm = T) %>% head()
mergeddata %>% group_by(year) %>% summarize_if(is.numeric,
   min, na.rm = T) %>% head()
mergeddata %>% group_by(year) %>% summarize_if(is.numeric,
    max, na.rm = T) %>% head()
```

### 4. Visualizing (30 pts)
```{r}
matrix <- mergeddata %>% na.omit() %>% select_if(is.numeric) %>% cor()
matrix

tidycor <- matrix %>% as.data.frame %>% rownames_to_column ("var1") %>% pivot_longer(-1, names_to="var2", values_to="correlation")

tidycor %>% ggplot(aes(var1, var2, fill = correlation)) + geom_tile() 
```


```{r}
ggplot(mergeddata, aes(`income per person`, `happiness scores`, color = `deaths by murder`)) + 
    geom_point(aes(size=`deaths by murder`)) + 
    coord_trans(x="log2") +
    ggtitle("Country Happiness Score vs Income per Person")+
    xlab("Income per Person ($)")+
    ylab("Happiness Score (0-.1)")+
    theme(axis.text.x = element_text(angle = 45))

```
The graph above maps three variables: a country's average happiness score, a countries average income per person, and a countries average death by murder rate. The X-axis represents income while the Y-axis represents the happiness score. Each data point was also sized and colored accordingly based on the number of deaths by murder. I found doing both made it easier to visualize since there were many data points present. An apparent trend portrayed in this graph is that that as the average income per person had increased, so had the country's happiness score.

```{r}
#income --> quartile income
mergeddata$incomequartiles <- cut_number(mergeddata$`income per person`, n=4, labels= c("1st quartile", "2nd quartile", "3rd quartile", "4th quartile"))

ggplot(mergeddata, aes(year,`deaths by murder`, fill = incomequartiles))+
    geom_bar(position = "dodge", stat="summary", fun=mean)+
    scale_y_continuous(name = "Death by Murder Rates", 
    breaks = c(0,2000,4000,6000,8000,9000,10000)) +
    xlab("Year")+
    ggtitle("Death by Murder Rates vs Year")
```
 The graph above maps three different variables: Death by Murder Rates, Years, and Income Quartiles. To view the other incomes better, I broke them down into quartiles and then added them into a separate column in my dataset. Although hard to tell, one apparent trend shown in the graph is that the countries with the highest income per person have the lowest death by murder rate every year.
 

### 5. Dimensionality Reduction (30 pts)

```{r}
        #1. Prepare Data
normdata <- mergeddata %>% select_if(is.numeric) %>% na.omit %>% scale
        #2. PCA (running princomp)
pca <- princomp(normdata)
summary(pca, loadings = T)
        #3. Choose number of principal components
eigenval <- pca$sdev^2
varprop = round(eigenval/sum(eigenval), 2)

ggplot() + geom_bar(aes(y = varprop, x = 1:3), stat = "identity") + 
    xlab("") + geom_path(aes(y = varprop, x = 1:3)) + geom_text(aes(x = 1:3, 
    y = varprop, label = round(varprop, 2)), vjust = 1, col = "white", 
    size = 5) + scale_y_continuous(breaks = seq(0, .6, .2), 
    labels = scales::percent) + scale_x_continuous(breaks = 1:3)
   
        #4. Plotting PC scores
dataframe <- data.frame(PC1 = pca$scores[, 1], PC2 = pca$scores[, 
    2])
ggplot(dataframe, aes(PC1, PC2)) + geom_point()+ ggtitle("PCA")+
    xlab("PC1")+
    ylab("PC2")


```

The two PC components used were Component 1 and Component 2 since their cumulative proportion of variance was greater than 80%. Looking at comp.1 and the graph, the data points that are high on PC1 tend to have high happiness scores and income per person. Mean while, the datapoints that are high on PC1 tend to have high death by murder rates and low happiness scores. Looking at the graph again PC1 also had higher variation than PC2 since most of the data points on the PC2 axis are below 0. 









